\begin{abstract}
We introduce \emph{predictive phase continuation}: effective variables are
calibrated at solvable nodes and used to predict, without refitting, phase
boundaries at held-out deformations.  The test is deliberately stronger than
curve similarity.  It requires semantic retention of the order parameter,
largest-size prediction against continuous, first-order-like, and crossover
alternatives, and a preregistered intervention near the predicted critical
window.  Our controlled four-domain benchmark separates outer disorder from
inner stochasticity and concentrates additional seeds only near inferred
boundaries.  \ifnum\ResultsReady=1 Across \PredictiveRuns{} registered tasks,
the median held-out boundary error falls from \BaseMedianError{} to
\AugmentedMedianError{} after adding a domain-specific macrovariable, a
\BridgeImprovement{} improvement.  Critical-window allocation removes
\MeanComputeSaving{} of uniform compute while preserving the registered
quality target.  \else Numerical claims are intentionally withheld until all
registered holdouts are complete. \fi Transformer and diffusion are the deep
controlled cases; reinforcement learning and multi-agent dynamics test the
same predictive grammar without assuming common equilibrium exponents.  An
artifact audit also identifies why saturated legacy order parameters cannot
support boundary claims.  The present evidence establishes a controlled
benchmark, not yet external validity on natural-data or open-weight endpoints.
We additionally organize the empirical Transformer study as a nine-axis tensor
over data, attention, MLP, residual parameterization, objective, optimizer,
scaling, lifecycle, and population. The completed five-seed aggregate covers
the controlled MLP, optimizer, objective, scaling, and bounded natural-data
edges; untested lifecycle and population cells remain explicit N0 entries in the
coverage matrix. The next aggregate records separate $O(1)$ training, IID-test,
and OOD risks, their generalization gaps, and learning trajectories, rather than
calling a held-out loss a generalization error. Existing results show that gated
MLPs need not improve an attention-dominant task, Muon and AdamW remain unresolved
at five seeds while exhibiting different gradient geometry from momentum SGD, and
the paths $N\propto d$ and $N\propto d^{3/2},d^2$ move held-out risk in opposite
directions. Byte-level TinyStories, SimpleStories, FineWeb-Edu, and Dolma establish
transport of the measurement protocol without asserting a universal boundary.
\end{abstract}
