\section{Introduction}

Large learning systems display abrupt changes in capability, representation, routing,
generation, and collective behavior. Calling every sharp curve a phase transition is,
however, scientifically weak. A useful statistical-mechanical claim must identify a
thermodynamic limit, an order parameter, a source of fluctuations, a control variable,
finite-size evidence, and a competing crossover explanation. It must also survive the
difference between quenched randomness in data or tasks and annealed randomness in
optimization or sampling.

This paper turns those requirements into a numerical program. Its core question is:
\begin{quote}
Can an order parameter and its transition be followed from a solvable anchor to a realistic
model through a sequence of controlled deformations, without changing the measured object
midway?
\end{quote}
We call this procedure \emph{phase continuation}. It resembles homotopy continuation:
an exact or oracle endpoint supplies semantics for the order parameter; adjacent models
relax one assumption; and a claim is weakened or rejected when the observable ceases to
track the same latent structure.

The first domain is the Transformer. Attention architectures
\citep{vaswani2017attention} combine several potentially distinct thermodynamic levels:
non-equilibrium training, internal representation and circuit organization, and output
sampling. We therefore use separate symbols and interventions for optimizer noise,
attention softmax scale, and generation temperature. We construct an architecture ladder
from tied rank-one attention to autoregressive generation, and a data ladder from Gaussian
tokens to hidden-state and grammar sources. The ladder supports tests of positional versus
semantic recovery, head specialization, attention--MLP division of labor, in-context
learning, long-context propagation, low-rank adaptation, glassy optimization, optimizer
universality, state tracking, expert routing, retrieval, multimodal alignment, and
compression.

The same logic is then applied to three additional domains. For diffusion, a Gaussian
mixture gives an exact score before locality, finite data, guidance, and reverse-trajectory
commitment are introduced. For reinforcement learning, a finite MDP gives an exact soft
policy before proxy rewards, rollout limits, preference estimation, and optimization
dynamics are introduced. For collective reasoning, mean-field population dynamics give a
controlled endpoint before sparse networks, heterogeneous influence, debate, committed
minorities, and role specialization are introduced. Four matched-latent experiments test
interactions among these domains rather than relying on metaphor.

Our contributions are:
\begin{enumerate}
\item A common phase-continuation formalism with an explicit assumption graph and matched
latent ensembles.
\item A Transformer M0--M8 and data D0--D5 experiment atlas covering programs A--N in one
manuscript.
\item Domain-specific diffusion, RL, and multi-agent phase cubes plus four cross-domain
continuations.
\item A strict evidence protocol using \NSeeds{} seeds, nested uncertainty, finite-size
scaling, rate functions, response, hysteresis, and evidence grades.
\item A portable implementation whose coverage audit fails when a proposed experiment is
missing and whose paper cannot silently report incomplete results.
\end{enumerate}
