\section{Related work}
\label{sec:related}

\paragraph{High-dimensional learning theory.}
Teacher--student models, replica calculations, approximate message passing,
and dynamical mean-field theory reduce high-dimensional learning to a small
set of overlaps and response functions \citep{gardner1989three,engel2001statistical,zdeborova2016statistical}.
Finite-size scaling then connects numerical systems to an asymptotic critical
point \citep{fisher1972scaling,binder1981finite,barber1983finite,privman1984universal}.
We import this evidentiary standard, but do not assume that a generic
Transformer has an equilibrium free energy or that every sharp training event
has a thermodynamic limit.

\paragraph{Solvable attention models.}
The direct foundation is the positional--semantic transition of
\citet{cui2024phase}, whose released code we use as an independent fidelity
oracle \citep{cui2024phasecode}.  Deep tied low-rank attention and sequence
multi-index models exhibit Bayes/AMP thresholds and sequential layer learning
\citep{troiani2025fundamental}; attention-indexed models extend the
Bayes-optimal perspective \citep{boncoraglio2025bayes}.  These results concern
stylized tied or otherwise structured operators.  In particular, the
attention-indexed result is not a published theory of QK spectral outliers;
spectral precedence is a hypothesis tested here, not a fact attributed to
that work.  Recent analyses of multi-head specialization
\citep{sagitova2026specialization}, LoRA in a solvable attention model
\citep{duranthon2026lora}, and chain-of-thought state tracking
\citep{forner2026learning} are informative but, as of this writing, the cited
versions are preprints and their assumptions delimit our comparisons.

\paragraph{Mechanisms and algorithmic phases.}
Linear attention admits asymptotic accounts of in-context regression
\citep{lu2025asymptotic}.  Markov-mixture training supports competing
retrieval and inference algorithms \citep{park2025competition}, while small
Transformers can implement distinct counting circuits
\citep{behrens2025counting}.  Mean-field token dynamics through attention
depth concerns forward representation evolution rather than learned-weight
dynamics \citep{rigollet2025meanfield}; we use it to motivate representation
observables, not as a training theory.

\paragraph{Optimization, scaling, and candidate transitions.}
$\mu$P stabilizes many width-dependent hyperparameters and motivates
declaring a scaling parameterization \citep{yang2021tensorprograms}.
Blockwise Hessian heterogeneity helps explain why Adam can outperform a single
global-rate SGD update in Transformers \citep{zhang2024adam}; it does not by
itself establish optimizer ``thermodynamic non-equivalence.''  We therefore
treat optimizer dependence as a robustness control.  Neural thermodynamic
laws have been proposed under a river--valley fast--slow approximation, with
limited GPT-2-small consistency checks \citep{liu2025neural}; our claims do not
inherit those assumptions.  Output-distribution distances can locate abrupt
changes \citep{arnold2024phase}, but a Jensen--Shannon peak is only a
candidate locator until order parameters and finite-size tests agree.
Likewise, discontinuous evaluation metrics can manufacture apparent
emergence \citep{schaeffer2023mirage}; all primary observables here are
continuous and their raw distributions are retained.

