\section{A hierarchical taxonomy of theory-to-reality deformations}
\label{sec:taxonomy}

The design space is a graph rather than a fully observed Cartesian tensor. A
node is a learning system; an edge removes one registered assumption. Three
levels prevent axes of different abstraction from being conflated.

\paragraph{Level 0: domain.}
The domains are Transformer representations, diffusion trajectories, RL
feedback dynamics, and multi-agent populations.

\paragraph{Level 1: thermodynamic level.}
The object of interest is respectively learning, representation, trajectory, or
population order. A system may occupy more than one level, but every PhaseCard
selects one primary level.

\paragraph{Level 2: six realism axes.}
Table~\ref{tab:taxonomy} is the central design taxonomy. Architecture
subcomponents such as attention and MLPs live under model/interaction rather
than appearing beside data or population as peer axes.

\begin{table}[t]
\centering\small
\begin{tabularx}{\linewidth}{@{}p{25mm}YYYY@{}}
\toprule
Axis & Transformer & Diffusion & RL / RLVR & Multi-agent \\
\midrule
Data / environment & synthetic, grammar, web text & mixture, manifold, image
data & bandit, MDP, verifier tasks & private evidence, task graph \\
Model / interaction & QKV, heads, MLP, norm, residual & score model, U-Net,
DiT, locality & tabular, MLP, Transformer policy & topology, bandwidth,
memory, roles \\
Objective / feedback & MSE, CE, causal, SFT, preference & $\epsilon$, $x_0$,
$v$, flow, guidance & gold, proxy, process reward & individual, team,
competitive reward \\
Dynamics / optimizer & SGD-M, AdamW, Muon, SOAP, schedule & optimizer, noise
schedule, solver & PPO, off-policy reuse, search & response and memory update \\
Scale / resources & $d,L,T,H,n,C$ & dimension, resolution, NFE & parameters,
horizon, rollouts & agents, degree, rounds \\
Lifecycle / population & pretrain, LoRA, pruning, RAG, rollouts &
distillation, conditioning & SFT--RL, verifier calls & heterogeneous agents,
dynamic topology \\
\bottomrule
\end{tabularx}
\caption{Hierarchical realism taxonomy. Entries are domain-specific subaxes,
not claims that every cell is tested in this paper.}
\label{tab:taxonomy}
\end{table}

\subsection{What is measured}

An order parameter must retain meaning under deformation. We group observables
by function rather than by domain (Table~\ref{tab:orders}). In particular,
$K_{\rm eff}^{\rm macro}=\exp H(Z)$ counts effective macrostates,
$N_{\rm eff}^{\rm corr}$ discounts correlated components, and
$r_{\rm eff}^{\rm spec}=(\sum_i\lambda_i)^2/\sum_i\lambda_i^2$ is a
spectral effective rank. They are not interchangeable effective numbers.

\begin{table}[t]
\centering\small
\begin{tabularx}{\linewidth}{@{}p{27mm}YYp{31mm}@{}}
\toprule
Family & Meaning & Representative probe & Required null \\
\midrule
Recovery & latent truth recovered & counterfactual semantic or oracle probe &
shuffled latent labels \\
Specialization & permutation symmetry broken & head, branch, strategy, or role
assignment & permuted components \\
Locality & range of causal influence & matched site or message perturbation &
random-site perturbation \\
Multiplicity & independent computational units & correlation participation &
independent-replica null \\
Entropy / commitment & macrostate selection & strategy or trajectory entropy &
token-level entropy alone \\
Replica structure & diversity of solutions & functional overlap distribution &
seed-matched random solutions \\
Robustness & quality of the ordered state & OOD risk, gold reward, truth
consensus & proxy-only performance \\
Resource response & marginal value of compute & matched FLOP, NFE, rollout, or
message intervention & equal-resource control \\
\bottomrule
\end{tabularx}
\caption{Order-parameter dictionary. A PhaseCard records the exact estimator,
probe distribution, normalization, units, and null.}
\label{tab:orders}
\end{table}

\subsection{Outcomes and PhaseCards}

An edge can preserve a phase structure, renormalize its coordinate, split or
merge regimes, round a sharp response, insert a new regime, separate
information and computation thresholds, become path dependent, or destroy the
semantics of the order parameter. These outcomes are hypotheses adjudicated by
registered diagnostics; they are not labels assigned from visual smoothness.

The minimal experimental record is
\begin{equation}
\mathcal P=(D,L,A,\delta,\omega,\mathbf N,\mathbf g,\mathbf m,
\mathbf R,H,E),
\end{equation}
where $D$ is domain, $L$ thermodynamic level, $A$ anchor, $\delta$ deformation,
$\omega$ disorder hierarchy, $\mathbf N$ system-size vector, $\mathbf g$
controls, $\mathbf m$ primary and secondary orders, $\mathbf R$ resources,
$H$ holdout rule, and $E$ evidence tier. Full PhaseCards are machine-readable;
representative cards appear in Section~\ref{sec:protocol}.
