\section{Phase-continuation framework}

\subsection{Assumption graph and continuation path}

A node in the assumption graph is
\[
  \mathcal{M}=(\mathcal{A},\mathcal{D},\mathcal{O},\mathcal{Q},\mathcal{U}),
\]
where \(\mathcal{A}\) is architecture or algorithm, \(\mathcal{D}\) the quenched
ensemble, \(\mathcal{O}\) the objective, \(\mathcal{Q}\) the dynamics or inference
rule, and \(\mathcal{U}\) the observable set. A directed edge changes one registered
assumption. Continuation is supported when (i) the oracle semantics of \(\mathcal{U}\)
are retained, (ii) adjacent distributions remain distinguishable but overlapping, and
(iii) the inferred phase boundary moves continuously within uncertainty. Failure of any
condition is recorded as a broken edge rather than hidden by redefining the order parameter.

Every domain begins with an exact or finite oracle. Approximation layers are then added:
finite samples, restricted interactions, learned scores or rewards, stochastic optimization,
and finally system components. The code records the family and variant in a content-addressed
\phasecard{} so two temperatures or two disorder ensembles cannot be conflated.

\subsection{Common coordinates}

For signed replica samples \(m_a\), system size \(N\), and macrostate probabilities
\(p_k\), the core observables are
\begin{align}
  \Order &= \mathbb{E}|m_a|, &
  \Sus &= N\left(\mathbb{E}m_a^2-(\mathbb{E}m_a)^2\right),\\
  \Binder &= 1-\frac{\mathbb{E}m_a^4}{3(\mathbb{E}m_a^2)^2}, &
  \Eff &= \exp\left[-\sum_k p_k\log p_k\right].
\end{align}
The macrostate entropy is \(S_{\rm macro}=\log \Eff\), while replica overlap is
\(\Replica_{ab}=N^{-1}\sum_i z_i^{(a)}z_i^{(b)}\). Interaction range is defined
operationally by a locality or ablation response. Generalization response is a finite
difference of \(\Gen\) under a registered distribution shift. Information load is the
number of independent latent constraints divided by the active degrees of freedom. These
coordinates have domain-specific estimators but common semantics.

A histogram \(P_N(m)\) gives the empirical rate function
\[
  \Rate_N(m)=-N^{-1}\log P_N(m)+C_N.
\]
We retain the samples required to recompute \(P_N\), multimodality, and barriers rather
than storing only means.

\subsection{Three Transformer thermodynamic levels}

We separate:
\begin{enumerate}
\item \emph{Training dynamics}: learning rate, gradient noise, Hessian spectrum, relaxation,
and an effective fluctuation--response temperature.
\item \emph{Representation and circuits}: attention scale, head and subspace overlaps,
QK/OV spectra, representation rank, and intervention response.
\item \emph{Generation}: sampling temperature, entropy, diversity, trajectory commitment,
and memorization.
\end{enumerate}
No numerical axis is called simply temperature in the manifest.

\subsection{Transformer and data ladders}

The architecture ladder is M0 tied rank-one; M1 untied Q/K; M2 QKVO; M3 multi-head; M4
residual and normalization; M5 MLP; M6 causal masking and rotary position; M7 depth; and M8
autoregressive generation. M0 forces teacher rank one even when a shared configuration
contains a higher rank. The data ladder is D0 Gaussian, D1 elliptical, D2 heavy-tailed,
D3 discrete codebook, D4 hidden Markov, and D5 grammar generated. Sample scaling follows
\(T=c d^p\), with \(p\in\{1,1.25,1.5,1.75,2\}\), while context, heads, depth,
vocabulary, rank, optimizer, and compute-fixed paths are separately registered.

The principal Transformer orders are teacher--student, functional positional and semantic,
head specialization, attention geometry, representation rank and anisotropy, algorithmic
state, in-context task recovery, QK/OV spectral outliers, replica overlap, and lifecycle
subspace drift.

\subsection{Domain-specific phase cubes}

For diffusion the axes are score approximation, noise/guidance, locality, sample support,
and reverse-time algorithm. Orders include semantic speciation, posterior multiplicity,
score nonlocality, memorization distance, trajectory commitment, and trajectory overlap.
For RL the axes are oracle versus proxy reward, policy entropy, rollout budget, optimizer,
and preference load. Orders include true/proxy return, occupancy overlap, entropy rate,
rollout multiplicity, and Goodhart gap. For multi-agent systems the axes are graph,
communication protocol, reliability heterogeneity, committed fraction, and population
size. Orders include consensus, polarization, correctness, influence concentration,
effective agent count, error reproduction number, and role specialization.
