\section{Numerical and statistical protocol}

\subsection{Seed hierarchy and estimand}

Every confirmatory condition uses exactly \NSeeds{} distinct outer seeds. An outer seed
defines quenched disorder. SHA-256 namespaced derivation produces data, initialization,
minibatch, dropout, and evaluation seeds. Inner stochastic replicates are averaged within
an outer seed before uncertainty is computed across outer seeds. Thus additional
checkpoints or minibatches cannot be counted as independent disorder realizations.

For outer estimates \(x_s\), the plotted interval is
\[
 \bar x \pm t_{0.975,4}\frac{s_x}{\sqrt{5}},\qquad t_{0.975,4}=2.776445.
\]
The manifest rejects four or six seeds. Aggregation rejects a missing run unless a
diagnostic command explicitly requests an incomplete report.

\subsection{Finite-size and critical-window protocol}

Exploratory scans locate candidate windows using adjacent Jensen--Shannon divergence,
Fisher sensitivity, and change-point scores. Confirmatory controls are fixed before the
five outer replicas are aggregated. Each transition claim uses at least six sizes and
reports: order parameter, \(\Sus\), \(\Binder\), \(P_N(m)\), rate function, peak drift,
and relaxation time. Up/down scans test hysteresis. A leading correction
\[
  m(N,t)=N^{-\beta/\nu}
  \left[f(tN^{1/\nu})+N^{-\omega}g(tN^{1/\nu})\right]
\]
is compared with a smooth crossover model. We report both fits and do not infer an
asymptotic singularity from one crossing.

\subsection{Response and non-equilibrium diagnostics}

Registered interventions include head or MLP ablation, attention-range truncation, guidance
change, reward/proxy perturbation, communication randomization, and committed-minority
injection. Response is paired within the same quenched realization. Optimizer runs retain
loss, overlap, gradient norm, Hessian proxies, and autocorrelation. A fluctuation--response
ratio is called an effective temperature only when its estimator and timescale are stated.

\subsection{Evidence grades}

Grade A requires all five outer seeds, at least six sizes, susceptibility growth, compatible
Binder behavior, a finite-size correction comparison, hysteresis testing, and nested
uncertainty. Grade B allows one missing asymptotic diagnostic but requires independent
order and response evidence. Grade C is a reproducible finite-size phenomenology. Anything
weaker is insufficient, not a transition. Cross-domain similarity does not imply a shared
universality class; exponents are compared only after separate within-domain grades.

\subsection{Programs A--N}

\begin{table}[t]
\centering
\small
\begin{tabularx}{\linewidth}{@{}lYY@{}}
\toprule
Program & Controlled deformation & Primary diagnostics \\
\midrule
A & M0--M8, D0--D5, positional/semantic mixture & \(m_{\rm pos},m_{\rm sem},\Sus,\Binder\) \\
B & head count to latent rank and tying & specialization, redundancy, effective heads \\
C & attention--MLP task and ablation ladder & functional error and paired intervention \\
D & finite/infinite task-pool ICL & context response and ridge oracle gap \\
E & context length, depth, rotary position & correlation length and slowing \\
F & LoRA rank continuation & subspace recovery and principal angles \\
G & random starts and energy landscape & \(P(q)\), barriers, relaxation \\
H & SGD, momentum, Adam, AdamW & Hessian/noise and effective temperature \\
I & Gaussian to grammar data & in/OOD error, covariance rank and anisotropy \\
J & direct answer to state tracking & scratchpad gain and commitment \\
K & top-k expert routing & load entropy, utilization, routing specialization \\
L & parametric to retrieval memory & memory load and paired error \\
M & paired latent modalities & canonical spectrum and collapse \\
N & quantization, pruning, distillation & functional fidelity and compression pressure \\
\bottomrule
\end{tabularx}
\caption{The unified Transformer program. Each row has a five-seed configuration and a
machine-readable falsifier.}
\end{table}

\subsection{Matched-latent cross-domain tests}

Four experiments reuse the same latent utility landscape: diffusion proposals followed by
language/RL selection; diffusion over policy logits followed by RL evaluation; multi-agent
belief aggregation followed by action selection; and MoE routing interpreted as influence
allocation among agents. Matching the latent instance permits paired oracle-gap and replica
overlap comparisons. The purpose is to test continuation coordinates, not to erase
domain-specific dynamics.
