\section{Experimental and statistical protocol}
\label{sec:protocol}

Evidence strength follows Table~\ref{tab:evidence}. The controlled predictive
benchmark uses 16 independent outer-disorder seeds, four inner stochastic
replicates, and 16 additional outer seeds near the frozen boundary window.
Inner replicates are never counted as independent disorder. The broad
Transformer taxonomy uses exactly five full-pipeline seeds per condition and
Student-$t$ intervals; it is a screen. Frozen confirmation uses twelve fresh
outer seeds. These designs answer different questions and are never pooled.

\begin{table}[t]
\centering\small
\begin{tabularx}{\linewidth}{@{}p{12mm}p{31mm}p{19mm}p{20mm}Y@{}}
\toprule
Tier & Purpose & Sizes & Outer seeds & Allowed language \\
\midrule
A & finite-size phase or boundary transport & $\geq6$ & $12$--$32$ &
phase structure, crossing/crossover, predictive boundary \\
B & frozen mechanism or endpoint comparison & $\geq3$ & $\geq5$ &
shift, mechanism candidate, resolved/unresolved contrast \\
C & natural or open endpoint & $\geq2$ & $\geq5$ &
measurement transport and bounded external validity \\
\bottomrule
\end{tabularx}
\caption{Evidence tiers. More seeds do not upgrade an invalid order parameter
or a non-independent holdout.}
\label{tab:evidence}
\end{table}

\paragraph{Finite-size inference.}
Finite-size scaling is domain specific \cite{fisher1972scaling,binder1981cumulants,
fisher1982firstorder}. We therefore do not impose $N^{-1/2}$ or $N^{-1}$ on
width, trajectory resolution, horizon, and population as universal exponents.
Candidate forms are empirical models whose parameters are fit below the
largest size and scored prospectively on it. A phase label additionally
requires order semantics, sharpening or crossing diagnostics, sensitivity to
the minimum size and fitting window, and a registered alternative.

\paragraph{Intensive observables.}
For a vocabulary of size $V$, cross entropy is reported as
$R_{\rm CE}=\ell_{\rm CE}/\log V$; the Brier risk is divided by
$1-1/V$. We report $R_{\rm train}$, $R_{\rm IID}$, $R_{\rm OOD}$, and
$e_{\rm gen}=R_{\rm IID}-R_{\rm train}$ separately. FLOPs, tokens, wall
time, width, depth, context, and sample count remain resource coordinates.
Causal module effects use raw risk differences and a bounded signed
normalization, never division by a vanishing remaining-accuracy scale.

\paragraph{Accounting and preregistration.}
A condition is a unique PhaseCard; a trained model is one outer seed; an inner
replicate is conditional stochasticity; an evaluation is not a new model.
Configuration hashes, corpus revisions and hashes, seed values, raw seed
metrics, censor flags, cap-hit flags, tokens, estimated FLOPs, and elapsed time
are stored before aggregation. Confirmation configurations were frozen after
the exploratory audit in the machine-readable claim ledger
\cite{nosek2018preregistration}.
