\section{Discussion and limitations}
\label{sec:discussion}

\paragraph{What a positive result would mean.}
A stable boundary through several ladder stages would not prove universality
for language models.  It would show that a particular competition between
positional displacement and latent semantic recovery survives named
architectural interventions along named scaling paths.  That is a useful,
mechanism-level statement and a sharper target for theory than an analogy to
thermodynamics in general.

\paragraph{What a negative result would mean.}
Disappearance is equally informative.  Because stages are cumulative and
instrumented identically, the first stage that loses functional order or
finite-size evidence locates the assumption responsible for the solvable
model's behavior.  A spectral outlier without semantic function, for example,
would falsify spectral norm as a sufficient mechanism; functional order
without a stable weight signature would instead favor reparameterization-
invariant descriptions.

\paragraph{Optimization is not a nuisance variable.}
SGD is part of the fidelity gate, while Adam/AdamW and momentum are robustness
interventions.  Curvature heterogeneity makes optimizer dependence plausible
\citep{zhang2024adam}, but different outcomes do not automatically constitute
distinct thermodynamic ensembles.  We report optimizer-resolved phase cards,
gradient/update norms, and convergence failures, and require loss-matched
comparisons before interpreting a boundary shift.

\paragraph{Finite size and nonequilibrium training.}
Our networks are finite, training is history dependent, and $M8$ is explicitly
nonequilibrium.  Binder-like statistics and rate functions are diagnostics of
replica distributions, not assertions of equilibrium Gibbs sampling.
Critical exponents are reported only when stable across width windows and
scaling choices.  Otherwise we use ``sharp crossover'' or ``candidate
transition.''

\paragraph{Synthetic data scope.}
D0--D5 isolate controlled departures from Gaussianity but are not natural
language.  The grammar is fixed length and the HMM has a small latent state;
neither captures open-vocabulary semantics.  Their purpose is causal
diagnosis: they reveal whether token correlation, discrete support, heavy
tails, or hierarchical latent structure is sufficient to change the phase
card.  Natural corpora are a later rung, not evidence retroactively attributed
to these ensembles.

\paragraph{Compute and multiplicity.}
Architecture, data, scale, optimizer, initialization, and seed create many
comparisons.  We distinguish a small set of registered headline hypotheses
from exploratory diagnostics, retain all state points, and do not select a
favorable width window after inspection.  Adaptive refinement creates a new
manifest.  The price is substantial compute; the benefit is that negative and
failed results remain auditable.

\paragraph{Reproducibility boundary.}
The repository records software and hardware provenance but cannot guarantee
bitwise identity across ROCm/PyTorch kernels.  Deterministic algorithms are
requested where available and scientific conclusions are based on
hierarchical variation, not one bitwise trajectory.  Raw model weights are not
required for paper reproduction; compact teacher states, trajectories,
diagnostics, and checksums are retained.

