\section{Limitations}

The new experiment is a controlled Tier-B stochastic benchmark.  It does not
establish natural-language, natural-image, open RLVR, or open-weight LLM-agent
external validity.  The previous repository contains trainable Tier-B+ smoke
models, but their saturated primary order failed the present audit and is not
promoted to evidence.  External endpoints require version-pinned models,
natural-data holdouts, matched token/FLOP controls, and independent
implementations.

The current empirical Transformer tensor does not yet cover sparse/MoE MLPs,
low-rank parameterizations, preference or RL objectives, open checkpoint
lifecycle trajectories, or agent populations. Those cells are displayed as N0
in the coverage matrix. The new train/test/OOD and mechanism measurements require
a fresh five-seed aggregate; legacy artifacts are retained as historical results
and are never relabeled with statistics they did not record.

Our finite sizes can distinguish competing predictive trends but do not by
themselves prove an asymptotic universality class.  RL and agent systems are
driven and potentially nonstationary; thermodynamic language is used only for
operational dynamic regimes.  Critical-window quality is benchmark-specific,
and compute savings do not transfer automatically to wall-clock savings on a
different architecture.

Finally, an adaptive design can bias inference if the first-stage boundary is
used to redefine the estimand.  We avoid this by freezing the original grid,
adding seeds rather than moving controls, and retaining both stages in the
manifest.
