\section{Reproducibility appendix}

\subsection{Repository structure}

Reusable code contains no site-specific absolute path. Immutable schemas, seed derivation,
metrics, and the runner registry live under \texttt{statphys/continuation/core}. Domain
kernels live under \texttt{domains}; finite-size, nested uncertainty, evidence, discovery,
and plotting live under \texttt{analysis}; Slurm and execution live under
\texttt{orchestration}. Root modules are compatibility facades.

The ten full configurations are:
\begin{enumerate}
\item \texttt{paper.toml}
\item \texttt{transformer\_atlas.toml}
\item \texttt{transformer\_systems.toml}
\item \texttt{domain\_programs.toml}
\item \texttt{cross\_domain.toml}
\item \texttt{learned\_transformer.toml}
\item \texttt{learned\_diffusion.toml}
\item \texttt{learned\_reinforcement.toml}
\item \texttt{learned\_multiagent.toml}
\item \texttt{continuation\_diagnostics.toml}
\end{enumerate}
Site-specific commands,
manifests, job IDs, and outputs are generated in the external works directory.

\subsection{Artifact contract}

Each task stores its complete specification, namespaced seeds, provenance, status, finite
metric JSON, compressed arrays, elapsed time, device, and checksums. A task ID hashes every
condition field except the outer seed and appends that seed. Atomic replacement prevents a
partially written file from appearing complete. Retry manifests contain all five replicas
of any incomplete condition, preserving the estimand.

\subsection{Coverage and falsification}

The coverage registry contains 58 runnable experiment IDs. Validation checks uniqueness,
implementation status, a registered family, at least one observable, a runner, a matching
configuration, and exactly five seeds in every configuration. Claims list their required
IDs, minimum evidence grade, and a sentence that would falsify the claim.

\subsection{Finite-size tables}

For every candidate boundary the released table contains size, control, five outer-seed
values, mean, standard deviation, standard error, 95 percent Student-t interval,
susceptibility, Binder cumulant, histogram, rate function, relaxation proxy, and intervention
response. Fit tables contain the raw scaling model, correction model, crossover model,
parameter uncertainty, and residual score. Exponent estimates are omitted when fewer than
five sizes pass artifact audit.

\subsection{Temperature ledger}

\begin{table}[h]
\centering
\begin{tabularx}{\linewidth}{@{}lYY@{}}
\toprule
Quantity & Operational definition & Forbidden conflation \\
\midrule
Training effective temperature & fluctuation--response ratio at stated timescale & learning rate alone \\
Attention temperature & divisor of attention logits & generation sampler temperature \\
Diffusion noise & forward corruption variance or schedule coordinate & policy entropy \\
Policy temperature & coefficient in entropy-regularized control & reward scale without normalization \\
Generation temperature & divisor of output logits & any training-dynamics quantity \\
\bottomrule
\end{tabularx}
\end{table}

\subsection{Minimal execution sequence}

First run the coverage command over all ten configurations. Expand each TOML file into an
immutable manifest. Render GPU or CPU Slurm arrays from a generic profile and supply
repository, manifest, output, and Python paths through environment variables. Monitor both
scheduler state and artifact state. Generate a retry manifest for any incomplete
condition. Only after strict aggregation should the plotting and paper-macro commands be
run, followed by the LaTeX build.
