\begin{abstract}
We introduce \emph{predictive phase continuation}: effective variables are
calibrated at solvable nodes and used to predict, without refitting, phase
boundaries at held-out deformations.  The test is deliberately stronger than
curve similarity.  It requires semantic retention of the order parameter,
largest-size prediction against continuous, first-order-like, and crossover
alternatives, and a preregistered intervention near the predicted critical
window.  Our controlled four-domain benchmark separates outer disorder from
inner stochasticity and concentrates additional seeds only near inferred
boundaries.  \ifnum\ResultsReady=1 Across \PredictiveRuns{} registered tasks,
the median held-out boundary error falls from \BaseMedianError{} to
\AugmentedMedianError{} after adding a domain-specific macrovariable, a
\BridgeImprovement{} improvement.  Critical-window allocation removes
\MeanComputeSaving{} of uniform compute while preserving the registered
quality target.  \else Numerical claims are intentionally withheld until all
registered holdouts are complete. \fi Transformer and diffusion are the deep
controlled cases; reinforcement learning and multi-agent dynamics test the
same predictive grammar without assuming common equilibrium exponents.  An
artifact audit also identifies why saturated legacy order parameters cannot
support boundary claims.  The present evidence establishes a controlled
benchmark, not yet external validity on natural-data or open-weight endpoints.
We additionally execute a 7,260-run, five-seed empirical tensor over MLP family,
optimizer, normalization, objective, scaling path, and natural-data source.
The resulting boundary transport is not reducible to parameter count: gated MLPs
need not improve an attention-dominant task, Muon and AdamW remain unresolved at
five seeds while exhibiting different gradient geometry from momentum SGD, and
the paths $N\propto d$ and $N\propto d^{3/2},d^2$ move held-out risk in opposite
directions.  Byte-level TinyStories, SimpleStories, FineWeb-Edu, and Dolma
experiments establish external transport of the measurement protocol without
asserting a universal numerical boundary.
\end{abstract}
